Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-10-12 paper.tex
changes in response to the referee

\begin{abstract}
  \C{We study the possibility for detecting gamma-ray emission from
  galaxy clusters. We consider 1) leptophilic models of dark matter
  (DM) annihilation that include a Sommerfeld enhancement (SFE), 2)
  different representative benchmark models of supersymmetric DM, and
  3) cosmic ray (CR) induced pion decay. To model DM annihilation, we
  consider continuum emission as a result of the hadronization of
  annihilating neutralinos, internal bremsstrahlung, and inverse
  Compton emission from the cosmic microwave background as well as
  from a realistic spatial and spectral distribution of dust and
  stellar light. Among all clusters/groups of a flux-limited X-ray
  sample, we predict Virgo, Fornax and M49 to be the brightest DM
  sources and find a particularly low CR-induced background for
  Fornax. For a minimum substructure mass given by the DM
  free-streaming scale, cluster halos maximize the substructure boost
  for which we find a factor of $\gtrsim 1000$. Since regions around
  the virial radius dominate the annihilation flux of substructures,
  the resulting surface brightness profiles are almost flat. This
  makes it very challenging to detect this flux with imaging
  atmospheric Cherenkov telescopes since their sensitivity drops
  approximately linearly with radius and they typically have $5-10$
  linear resolution elements across a cluster. Assuming cold dark
  matter with a substructure mass distribution down to an Earth mass
  and using extended \Fermi upper limits, we rule out the leptophilic
  models in their present form in 28 clusters, and limit the boost
  from SFE in M49 and Fornax to be $\lesssim 5$. This corresponds to a
  limit on SFE in the Milky Way of $\lesssim 3$ which is too small to
  account for the increasing positron fraction with energy as seen by
  PAMELA. Alternatively, if SFE is realized in Nature, this would
  imply a limiting substructure mass of $\mlim > 10^4\,\msun$---a
  problem for structure formation in most particle physics
  models. Using individual cluster observations, it will be
  challenging for \Fermi to constrain our selection of DM benchmark
  models without SFE. The \Fermi upper limits are, however, closing in
  on our predictions for the CR flux using an analytic model based on
  cosmological hydrodynamical cluster simulations. We limit the
  CR-to-thermal pressure in nearby bright galaxy clusters of the
  \Fermi sample to $\lesssim 10\%$ and in Norma and Coma to $\lesssim
  3\%$. Thus we will soon start to constrain the underlying CR physics
  such as shock acceleration efficiencies or CR transport properties.}
\end{abstract}

@@

\C{In this paper, we will investigate in detail the potential of
  several of the most promising galaxy clusters to produce an
  annihilation gamma-ray yield which could be observable with present
  and planned gamma-ray detectors. Here, we sketch out the main
  arguments that need to be considered for maximizing the expected
  signal-to-noise ratio of a promising target cluster. First, we need
  to maximize the DM annihilation flux of an unresolved cluster,
  $F=A_\rmn{dm} \int dV \rho^2 / D^2 \sim \mvir^\alpha/D^2$, where
  $A_\rmn{dm}$ depends on the particle physics model of DM, $\rho$ is
  the smooth DM density profile, $D$ is the luminosity distance, and
  $\alpha=1$ if we assume universality of the DM density profile. The
  dependence of $\rho$ on the halo formation epoch breaks the
  universality and slightly modifies the mass dependence, yielding
  $\alpha=0.83$ \cite{2009PhRvL.103r1302P}. Additionally, the presence
  of a hierarchy of substructures down to small scales
  \cite{2008Natur.456...73S} and the potential dependence of the
  particle physics cross section on the relative DM particle
  velocities \cite{ArkaniHamed:2008qn} may furthermore modify the
  scaling parameter $\alpha$ and shall be one of the focus points of
  this work. Second, we need to minimize the expected noise which is a
  sum of galactic and cluster-intrinsic foreground. While the galactic
  foreground varies across the sky, it is typically lower for
  increasing latitude (away from the galactic plane). It is thought
  that the gamma-ray foreground from clusters is dominated by CRs
  which are accelerated at cosmological formation shocks and
  transported over time although there could be a substantial
  contribution of CRs from AGNs and supernova-driven galactic
  winds. In this work we will also study (and constrain) the
  CR-induced emission from clusters to identify those objects that are
  expected to be especially dim which would imply a low CR-induced
  background.  Since the CR-induced emission is expected to scale with
  the thermal X-ray emission of clusters \cite{2004A&A...413...17P},
  we estimate gamma-ray fluxes of a nearby X-ray flux complete sample
  of galaxy clusters from the Rosat all-sky survey (extended HIghest
  X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
  \cite{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
    we have added the Virgo cluster to the sample that we nevertheless
    refer to as extended HIFLUGCS catalogue in the following.}  This
  complements and extends previous work related to DM in clusters
  \cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2011arXiv1110.1529H}
  and to CRs in clusters
  \cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.}

@@

\subsubsection{Leptophilic models}
\label{sect:LP}
\C{There is also a possibility to search for DM annihilation in the Milky
Way indirectly through annihilation to antimatter channels, which
however lacks the important directional signature of
gamma-rays. Recently, it has however been much in focus due to the
surprising findings of PAMELA \cite{Adriani:2008zr} and \Fermi
\cite{Abdo:2009zk}. Viable particle physics models are rather
constrained by other observations, however. For example, the
non-observation of an enhanced antiproton flux by PAMELA
\cite{Adriani:2010rc} means that quark-antiquark final states have to
be suppressed. This has led to the postulate of a whole class of
{\em leptophilic} models with mainly or exclusively annihilation to
leptons and antileptons. In addition, one has to have a sizeable
fraction to final states containing muons or tau leptons as the shape
of the spectrum disfavors direct annihilation to electrons and
positrons only. A large enhancement, of the order of at least several
hundred, of the annihilation rate is also needed, something that may
be given by the {\em Sommerfeld effect} (for a pedagogical review of
viable scenarios, see \cite{ArkaniHamed:2008qn}). As still another
difficulty for these models, the lack of an IC signature from the
Galactic center means that the DM density distribution must be cored
rather than cuspy
\cite{Bertone:2008xr,Cirelli:2008pk,Bergstrom:2008ag}. (For a review
of the DM modeling of these effects, see \cite{Bergstrom:2009ib}. For
a recent treatment, showing still viable models, see
\cite{Finkbeiner:2010sm}.)}

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of
($1/4:1/4:1/2$) into ($\mu^+\mu^-:\e^+\e^-:\pi^+\pi^-$). Furthermore,
since most of the gamma-ray flux is expected to come from dense
subhalos within clusters which have a velocity dispersion close to the
velocity limit $v_\rmn{sat}$ where the SFE saturates, we use
$S=\B_\sfe(v_\rmn{sat})\approx 530$ for the cluster
halos.\footnote{Note that $S$ is strictly speaking the Sommerfeld
  enhanced annihilation cross-section, but here we have additionally
  included the order unity constraint that ensures that the relic
  density is not over-depleted at freeze-out.} However, for figures
where we neglect the substructure boost, we adopt a velocity dependent
SFE that is normalized to fit the electron and positron excess observed at
Earth\del{observed by the PAMELA and \Fermi satellites},
\begin{equation}
\B_\sfe(\sigv) = \B_\rmn{sfe,MW} \frac{v_\rmn{MW}}{\sigv} =
70 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,,
\label{eq:B_sfe}
\end{equation}
\C{where the local boost factor required to explain the data
  $\B_\rmn{sfe,MW}=300$.} \del{Here the boost factor needed to explain
  the PAMELA/\Fermi excess of electrons and positrons
  $\B_\rmn{sfe,MW}=300$,} The velocity dispersion for the MW
$v_\rmn{MW} \approx 220\,\rmn{km\,s}^{-1}$, and the \C{mass dependent}
velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. Note that, clusters have a larger velocity
dispersion and an accordingly smaller Sommerfeld boost factor.

@@

We now derive the squared density profile for the substructures
using
\begin{eqnarray}
\rho_\rmn{sub}^2(r) &=& \frac{\dd L_\rmn{sub}}{\dd V} \frac{1}{A_{\rmn{dm}}}\,,\label{eq:rho_sub}\\
L_{200\sm} &=& A_{\rmn{dm}} \int_{\rvir} \dd r' 4\pi r'^2 \rho^2(r')\,,\label{eq:Lsm}\\
 \frac{\dd L_\rmn{sub}}{\dd V} &=& a_0\,a_1\,C(\mvir)\,x^{g(x)}\,
\left[\frac{1+a_2\log(x)}{4\pi\,r^3}\right]\,,\nonumber\\
\\
x &=& r/\rvir\,,\quad g(x) = a_2+a_1\,x^{a_2}\,.\label{eq:xvir}
\end{eqnarray}
Here, $A_{\rmn{dm}} = q_{\rmn{sm},i}(\eg,R) / \rho(R)^2$ represents
the particle physics factor, where $q_{\rmn{sm},i}(\eg,R)$ is defined
by Eq.~(\ref{eq:q_sm}). The different density profiles have some
impact on the luminosity from annihilating DM, although the details of
the density profile can be neglected compared to the dominating boost
from substructures (assuming DM to be cold). In
Fig.~\ref{fig:radial_lum} we compare the radial dependence of the
accumulative luminosity from different smooth cluster density profiles
to the boosted luminosity due to substructure for different mass
scales. We recalculate the overdensity, $\rhos$, for the shallower
density profile with $\beta=0.6$ and rescale the concentration
parameter in Eq.~(\ref{eq:cfit}) with $300/160$
\cite{2011ApJ...728L..39N} to account for the more centralized scale
radius in cluster with a shallow inner slope. The emission of this
profile with $\beta=0.6$ is about 30\% larger within $\rvir$ compared
to a cuspy NFW ($\beta=1.0$). This difference is built up within
$0.1\rvir$ (i.e. close to $r_\s$). Hence the slope of the central part
of a cluster has little influence for the DM luminosity within $\rvir$
as long as the degeneracy in the lensing mass measurements with $r_\s$
(which decreases for decreasing inner slope) has been taken into
account. The emission from an Einasto density profile is about 50\%
larger than the cuspy NFW profile in the periphery of the cluster,
where the difference is mainly built up at a few percent of
$\rvir$. The increase in luminosity due to substructures is negligible
in the center of halos, but integrated out to $\rvir$ the boost
relative to the smooth emission profile amounts to approximately 20
for dwarf galaxies, $200$ for galaxies, and $10^3$ for galaxy
clusters. \C{We stress that these boost factors are only realized in
  the region around the virial radius of each respective halo which
  are mostly stripped for dwarf substructures in the MW. Hence a more
  realistic boost from substructures is probably much smaller in Dwarf
  galaxies. In addition,} these boost factors are only realized for
direct annihilation emission (continuum emission or final state
radiation) or IC scattering off homogeneous seed photon fields
(CMB). For IC scattering of SD photons, the overlap of final state
leptons and SD photons is smaller which causes the substructure boost
over the smooth emission to be reduced by roughly two orders of
magnitudes. It should also be noted that the substructure boosted
fluxes from clusters are much more extended than for dwarf galaxies,
hence more difficult to detect with Cherenkov telescopes. In fact, it
was shown in Sanchez-Conde et al. \cite{2011arXiv1104.3530S} that
inside $0.1$~deg, where the sensitivity of Cherenkov telescopes is
maximized, the expected DM flux including substructures from the
brightest dwarf galaxy is about an order magnitude higher than the
brightest cluster. However, that work assumed a substructure boost of
about 50 in the most massive cluster halos which is more than an order
magnitude smaller than what we use in this work, hence in projection
we expect this difference to be a factor few smaller.

@@

In supernovae remnants and on scales of galaxies, there are convincing
evidences of non-thermal populations. Especially, in the MW, the
cosmic rays are observed directly as well as indirectly through radio,
X-ray, and gamma-ray emission. On larger scales of the order of few
100~kpc up to Mpc, there are currently a vast number of observations
of radio emission coming from radio mini halos in the center of
cooling flow clusters, radio relics in the periphery of clusters
\cite{2004rcfg.proc..335K} as well as smooth giant radio halos that
have been observed in more than 50 clusters
\cite{2003ASPC..301..143F,2008SSRv..134...93F}. This type of emission
is expected in clusters since the formation process of a galaxy
cluster is a very energetic processes that induces both turbulence as
well as frequently occurring merging and accretion shocks which both
are thought to accelerate relativistic non-thermal protons and radio
emitting electrons to high energies. Thus most of the observed radio
emission in clusters is expected to emerge from cosmic-ray electron
induced radio synchrotron emission. The precise origin of these
electrons, in especially relics and giant radio halos, is still not
settled. One possible scenario for the production of these electrons
are hadronic CR interactions with ambient gas protons which results in
the production of both charged and neutral pions that decay into
electrons, neutrinos, and gamma-rays (see
\cite{1980ApJ...239L..93D,1982AJ.....87.1266V,1999APh....12..169B,2000A&A...362..151D,2001ApJ...559...59M,2003MNRAS.342.1009M,2004A&A...413...17P,2004MNRAS.352...76P,2008MNRAS.385.1211P,2008MNRAS.385.1242P,2009JCAP...09..024K,2010MNRAS.401...47D,2010MNRAS.407.1565D,2010ApJ...722..737K,2011A&A...527A..99E}
).\footnote{An alternative scenario is the second order turbulent
  re-acceleration through the interaction of a previously injected
  relativistic population of electrons by supernova driven winds or
  active galactic nuclei with plasma waves and
  magneto-turbulence. \protect
  \cite{1987A&A...182...21S,1993ApJ...406..399G,2004MNRAS.350.1174B,2005MNRAS.363.1173B,2007MNRAS.378..245B,2011MNRAS.410..127B,2009A&A...507..661B}}
Supporting evidence comes from the smoothness of the extended radio
emission that often resembles that observed in thermal X-rays. This
can be easily achieve in the hadronic model since the long cooling
time of cosmic ray protons (CRs) allows for an extended and large
population of CRs to build up over the cluster history
\cite{1996SSRv...75..279V, 1997ApJ...477..560E,
  1997ApJ...487..529B}. The production of these secondaries depend
both on the gas and CR densities in the cluster, where the CR
density roughly traces the gas outside the core regime and is slightly
enhanced in the center. This density scaling implies that clusters
are great targets for Cherenkov telescopes with a high sensitivity for
the central parts of nearby clusters. Detecting the cluster gamma-ray
emission is crucial in this respect as it potentially provides the
unique and unambiguous evidence of CR populations in clusters through
observing the $\pi^0$ bump at about 100~MeV in the spectra.

We adopt the universal spectral and spatial gamma-ray model developed
by Pinzke \& Pfrommer \cite{2010MNRAS.409..449P} to estimate the
emission from decaying $\pi^0$:s that dominates over the IC emission
from primary and secondary electrons above 100~MeV in clusters. The
gamma-ray formalism was derived from high resolution simulations of
clusters of galaxies that included radiative hydrodynamics, star
formation, supernova feedback, and followed CR physics using a novel
formulation that trace the most important injection and loss processes
self-consistently while accounting for the CR pressure in the
equation of motion
\cite{2008A&A...481...33J,2007A&A...473...41E,2006MNRAS.367..113P}.
We note that the overall normalization of the CR and gamma-ray
distribution scales with the maximum acceleration efficiency at
structure formation shock waves. Following recent observations at
supernova remnants \cite{2009Sci...325..719H} as well as theoretical
studies \cite{2005ApJ...620...44K}, we adopt a realistic value of this
parameter and assume that 50\% of the dissipated energy at strong
shocks is injected into CRs while this efficiency rapidly decreases
for weaker shocks \cite{2007A&A...473...41E}. As a caveat we note that
the analytic model is based on simulations that only considered
advective transport of CRs by turbulent gas motions which produces
centrally enhanced profiles. However active CR transport such as CR
diffusion and streaming tends to drive the CR radial profiles towards
being flat, with equal CR number density everywhere. While the CR
streaming is then usually larger than typical advection velocities and
becomes comparable or lower than this only for periods with trans- and
supersonic cluster turbulence during a cluster merger. As a
consequence a bimodality of the CR spatial distribution results with
merging (relaxed) clusters showing a centrally concentrated (flat) CR
energy density profile \cite{2011A&A...527A..99E}. This translates
into a bimodality of the expected diffuse radio and gamma-ray emission
of clusters, since more centrally concentrated CR will find higher
target densities for hadronic CR proton interactions
\cite{2011A&A...527A..99E}. As a result of this, relaxed clusters
could have a reduced gamma-ray luminosity by up to a factor of five.

@@

We focus in this section on the spectral distribution of gamma-rays
from clusters. In the LP model, DM annihilation radiation includes
final state radiation and gives rise to substantial amounts of
electrons and positrons that IC upscatter background radiation fields
to high energies. We also consider four supersymmetric DM models with
a high gamma-ray yield in the form of continuum emission and IC
induced emission. In addition to the annihilating DM, we estimate the
gamma-ray flux induced by shock-accelerated CRs. \C{Note that all
  fluxes in this section are derived within an angle corresponding to
  $\rvir$ and are smoothed over using a PSF of $0.1\deg$.}

>>>>

\begin{abstract}
  \C{We study the possibility for detecting gamma-ray emission from galaxy
    clusters. We consider 1) leptophilic models of dark matter (DM) annihilation
    that include a Sommerfeld enhancement (SFE), 2) different representative
    benchmark models of supersymmetric DM, and 3) cosmic ray (CR) induced pion
    decay. Among all clusters/groups of a flux-limited X-ray sample, we predict
    Virgo, Fornax and M49 to be the brightest DM sources and find a particularly
    low CR-induced background for Fornax. For a minimum substructure mass given
    by the DM free-streaming scale, cluster halos maximize the substructure
    boost for which we find a factor of $\gtrsim 1000$. Since regions around the
    virial radius dominate the annihilation flux of substructures, the resulting
    surface brightness profiles are almost flat. This makes it very challenging
    to detect this flux with imaging atmospheric Cherenkov telescopes since
    their sensitivity drops approximately linearly with radius and they
    typically have $5-10$ linear resolution elements across a cluster. Assuming
    cold dark matter with a substructure mass distribution down to an Earth mass
    and using extended \Fermi upper limits, we rule out the leptophilic models
    in their present form in 28 clusters, and limit the boost from SFE in M49
    and Fornax to be $\lesssim 5$. This corresponds to a limit on SFE in the
    Milky Way of $\lesssim 3$ which is too small to account for the increasing
    positron fraction with energy as seen by PAMELA and challenges the DM
    interpretation. Alternatively, if SFE is realized in Nature, this would
    imply a limiting substructure mass of $\mlim > 10^4\,\msun$---a problem for
    structure formation in most particle physics models. Using individual
    cluster observations, it will be challenging for \Fermi to constrain our
    selection of DM benchmark models without SFE. The \Fermi upper limits are,
    however, closing in on our predictions for the CR flux using an analytic
    model based on cosmological hydrodynamical cluster simulations. We limit the
    CR-to-thermal pressure in nearby bright galaxy clusters of the \Fermi sample
    to $\lesssim 10\%$ and in Norma and Coma to $\lesssim 3\%$. Thus we will
    soon start to constrain the underlying CR physics such as shock acceleration
    efficiencies or CR transport properties.}
\end{abstract}

% To model DM annihilation, we
%   consider continuum emission as a result of the hadronization of
%   annihilating neutralinos, internal bremsstrahlung, and inverse
%   Compton emission from the cosmic microwave background as well as
%   from a realistic spatial and spectral distribution of dust and
%   stellar light.

@@

\C{In this paper, we will investigate in detail the potential of
  several of the most promising galaxy clusters to produce an
  annihilation gamma-ray yield which could be observable with present
  and planned gamma-ray detectors. Here, we sketch out the main
  arguments that need to be considered for maximizing the expected
  signal-to-noise ratio of a promising target cluster. First, we need
  to maximize the DM annihilation flux of an unresolved cluster,
  $F=A_\rmn{dm} \int dV \rho^2 / D^2 \propto \mvir^\alpha/D^2$, where
  $A_\rmn{dm}$ depends on the particle physics model of DM, $\rho$ is
  the smooth DM density profile, $D$ is the luminosity distance, and
  $\alpha=1$ if we assume universality of the DM density profile. The
  dependence of $\rho$ on the halo formation epoch breaks the
  universality and slightly modifies the mass dependence, yielding
  $\alpha=0.83$ \cite{2009PhRvL.103r1302P}. Additionally, the presence
  of a hierarchy of substructures down to small scales
  \cite{2008Natur.456...73S} and the potential dependence of the
  particle physics cross section on the relative DM particle
  velocities \cite{ArkaniHamed:2008qn} may furthermore modify the
  scaling parameter $\alpha$ and shall be one of the focus points of
  this work. Second, we need to minimize the expected noise which is a
  sum of instrumental noise, galactic and cluster-intrinsic
  foreground. While the galactic foreground varies across the sky, it
  is typically lower for increasing latitude (away from the galactic
  plane). It is thought that the gamma-ray foreground from clusters is
  dominated by CRs which are accelerated at cosmological formation
  shocks and transported over time although there could be a
  substantial contribution of CRs from AGNs and supernova-driven
  galactic winds. In this work we will also study (and constrain) the
  CR-induced emission from clusters to identify those objects that are
  expected to be especially dim which would imply a low CR-induced
  background.  Since the CR-induced emission is expected to scale with
  the thermal X-ray emission of clusters \cite{2004A&A...413...17P},
  we estimate gamma-ray fluxes of a nearby X-ray flux complete sample
  of galaxy clusters from the Rosat all-sky survey (extended HIghest
  X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
  \cite{2002ApJ...567..716R,2007A&A...466..805C}).\footnote{Note that
    we have added the Virgo cluster to the sample that we nevertheless
    refer to as extended HIFLUGCS catalogue in the following.}  This
  complements and extends previous work related to DM in clusters
  \cite{2006A&A...455...21C,2009PhRvD..80b3005J,2011arXiv1104.3530S,2011ApJ...726L...6C,2011arXiv1110.1529H}
  and to CRs in clusters
  \cite{2010MNRAS.409..449P,2011MNRAS.410..127B,2008MNRAS.385.1211P,2009JCAP...08..002K,2010MNRAS.407.1565D}.}

@@

\subsubsection{Leptophilic models}
\label{sect:LP}
\C{There is also a possibility to search for DM annihilation in the Milky
Way indirectly through annihilation to antimatter channels, which
however lacks the important directional signature of
gamma-rays. Recently, it has however been much in focus due to the
surprising findings of PAMELA \cite{Adriani:2008zr} and \Fermi
\cite{Abdo:2009zk}. Viable particle physics models are rather
constrained by other observations, however. For example, the
non-observation of an enhanced antiproton flux by PAMELA
\cite{Adriani:2010rc} means that quark-antiquark final states have to
be suppressed. This has led to the postulate of a whole class of
{\em leptophilic} models with mainly or exclusively annihilation to
leptons and antileptons. In addition, one has to have a sizeable
fraction of final states containing muons or tau leptons as the shape
of the spectrum disfavors direct annihilation to electrons and
positrons only. A large enhancement, of the order of at least several
hundred, of the annihilation rate is also needed, something that may
be given by the {\em Sommerfeld effect} (for a pedagogical review of
viable scenarios, see \cite{ArkaniHamed:2008qn}). As still another
difficulty for these models, the lack of an IC signature from the
Galactic center means that the DM density distribution must be cored
rather than cuspy
\cite{Bertone:2008xr,Cirelli:2008pk,Bergstrom:2008ag}. (For a review
of the DM modeling of these effects, see \cite{Bergstrom:2009ib}. For
a recent treatment, showing still viable models, see
\cite{Finkbeiner:2010sm}.)}

@@

We adopt our benchmark LP model from \cite{Finkbeiner:2010sm} where we
use a DM mass $m_\chi=1.6$~TeV and a branching ratio of
($1/4:1/4:1/2$) into ($\mu^+\mu^-:\e^+\e^-:\pi^+\pi^-$). Furthermore,
since most of the gamma-ray flux is expected to come from dense
subhalos within clusters which have a velocity dispersion close to the
velocity limit $v_\rmn{sat}$ where the SFE saturates, we use
$S=\B_\sfe(v_\rmn{sat})\approx 530$ for the cluster
halos.\footnote{Note that $S$ is strictly speaking the Sommerfeld
  enhanced annihilation cross-section, but here we have additionally
  included the order unity constraint that ensures that the relic
  density is not over-depleted at freeze-out.} However, for figures
where we neglect the substructure boost, we adopt a velocity dependent
SFE that is normalized to fit the electron and positron excess observed at
Earth\del{observed by the PAMELA and \Fermi satellites},
\begin{equation}
\B_\sfe(\sigv) = \B_\rmn{sfe,MW} \frac{v_\rmn{MW}}{\sigv} =
70 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,,
\label{eq:B_sfe}
\end{equation}
\C{where the local boost factor required to explain the data is
  $\B_\rmn{sfe,MW}=300$.} \del{Here the boost factor needed to explain
  the PAMELA/\Fermi excess of electrons and positrons
  $\B_\rmn{sfe,MW}=300$,} The velocity dispersion for the MW,
$v_\rmn{MW} \approx 220\,\rmn{km\,s}^{-1}$, and the \C{mass dependent}
velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. \del{Note that, clusters have a larger velocity
dispersion and an accordingly smaller Sommerfeld boost factor.}

@@

We now derive the squared density profile for the substructures
using
\begin{eqnarray}
\rho_\rmn{sub}^2(r) &=& \frac{\dd L_\rmn{sub}}{\dd V} \frac{1}{A_{\rmn{dm}}}\,,\label{eq:rho_sub}\\
L_{200\sm} &=& A_{\rmn{dm}} \int_{\rvir} \dd r' 4\pi r'^2 \rho^2(r')\,,\label{eq:Lsm}\\
 \frac{\dd L_\rmn{sub}}{\dd V} &=& a_0\,a_1\,C(\mvir)\,x^{g(x)}\,
\left[\frac{1+a_2\log(x)}{4\pi\,r^3}\right]\,,\nonumber\\
\\
x &=& r/\rvir\,,\quad g(x) = a_2+a_1\,x^{a_2}\,.\label{eq:xvir}
\end{eqnarray}
Here, $A_{\rmn{dm}} = q_{\rmn{sm},i}(\eg,R) / \rho(R)^2$ represents
the particle physics factor, where $q_{\rmn{sm},i}(\eg,R)$ is defined
by Eq.~(\ref{eq:q_sm}). The different density profiles have some
impact on the luminosity from annihilating DM, although the details of
the density profile can be neglected compared to the dominating boost
from substructures (assuming DM to be cold). In
Fig.~\ref{fig:radial_lum} we compare the radial dependence of the
accumulative luminosity from different smooth cluster density profiles
to the boosted luminosity due to substructure for different mass
scales. We recalculate the overdensity, $\rhos$, for the shallower
density profile with $\beta=0.6$ and rescale the concentration
parameter in Eq.~(\ref{eq:cfit}) with $300/160$
\cite{2011ApJ...728L..39N} to account for the more centralized scale
radius in cluster with a shallow inner slope. The emission of this
profile with $\beta=0.6$ is about 30\% larger within $\rvir$ compared
to a cuspy NFW ($\beta=1.0$). This difference is built up within
$0.1\rvir$ (i.e. close to $r_\s$). Hence the slope of the central part
of a cluster has little influence for the DM luminosity within $\rvir$
as long as the degeneracy in the lensing mass measurements with $r_\s$
(which decreases for decreasing inner slope) has been taken into
account. The emission from an Einasto density profile is about 50\%
larger than the cuspy NFW profile in the periphery of the cluster,
where the difference is mainly built up at a few percent of
$\rvir$. The increase in luminosity due to substructures is negligible
in the center of halos, but integrated out to $\rvir$ the boost
relative to the smooth emission profile amounts to approximately 20
for dwarf galaxies, $200$ for galaxies, and $10^3$ for galaxy
clusters. \C{We stress that these boost factors are only realized in
  the region around the virial radius of each respective halo which
  is mostly tidally stripped for dwarfs in the MW. Hence a more
  realistic boost from substructures is probably much smaller in
  satellite dwarf galaxies. In addition,} these boost factors are only
realized for direct annihilation emission (continuum emission or final
state radiation) or IC scattering off homogeneous seed photon fields
(CMB). For IC scattering of SD photons, the overlap of final state
leptons and SD photons is smaller which causes the substructure boost
over the smooth emission to be reduced by roughly two orders of
magnitudes. It should also be noted that the substructure boosted
fluxes from clusters are much more extended than for dwarf galaxies,
hence more difficult to detect with Cherenkov telescopes. In fact, it
was shown in Sanchez-Conde et al. \cite{2011arXiv1104.3530S} that
inside $0.1$~deg, where the sensitivity of Cherenkov telescopes is
maximized, the expected DM flux including substructures from the
brightest dwarf galaxy is about an order magnitude higher than the
brightest cluster. However, that work assumed a substructure boost of
about 50 in the most massive cluster halos which is more than an order
magnitude smaller than what we use in this work, hence in projection
we expect this difference to be a factor few smaller.

@@

In supernovae remnants and on scales of galaxies, there are convincing evidences
of non-thermal populations. Especially, in the MW, the cosmic rays are observed
directly as well as indirectly through radio, X-ray, and gamma-ray emission. On
larger scales of the order of few 100~kpc up to Mpc, there are currently a vast
number of observations of radio emission coming from radio mini halos in the
centers of cooling flow clusters, radio relics in the periphery of clusters
\cite{2004rcfg.proc..335K}, as well as giant radio halos amounting to a total
number of more than 50 diffuse radio sources in clusters
\cite{2003ASPC..301..143F,2008SSRv..134...93F}. This type of emission is
expected in clusters since the formation process of a galaxy cluster is a very
energetic processes that induces both turbulence as well as frequently occurring
merging and accretion shocks which both are thought to accelerate relativistic
non-thermal protons and radio emitting electrons to high energies. The precise
origin of these electrons, in especially relics and giant radio halos, is still
not settled. One possible scenario for the production of these electrons are
hadronic CR interactions with ambient gas protons which results in charged and
neutral pions that decay into electrons, neutrinos, and gamma-rays (see
\cite{1980ApJ...239L..93D,1982AJ.....87.1266V,1999APh....12..169B,2000A&A...362..151D,2001ApJ...559...59M,2003MNRAS.342.1009M,2004A&A...413...17P,2004MNRAS.352...76P,2008MNRAS.385.1211P,2008MNRAS.385.1242P,2009JCAP...09..024K,2010MNRAS.401...47D,2010MNRAS.407.1565D,2010ApJ...722..737K,2011A&A...527A..99E}
).\footnote{An alternative scenario is the second order turbulent
  re-acceleration through the interaction of a previously injected relativistic
  population of electrons by supernova driven winds or active galactic nuclei
  with plasma waves and magneto-turbulence. \protect
  \cite{1987A&A...182...21S,1993ApJ...406..399G,2004MNRAS.350.1174B,2005MNRAS.363.1173B,2007MNRAS.378..245B,2011MNRAS.410..127B,2009A&A...507..661B}}
Supporting evidence comes from the smoothness of the extended radio emission
that often resembles that observed in thermal X-rays. This can be easily
explained in the hadronic model since the long cooling time of cosmic ray
protons (CRs) of order the Hubble time allows for a cluster-filling population
of CRs to build up over the formation history \cite{1996SSRv...75..279V,
  1997ApJ...477..560E, 1997ApJ...487..529B}. The production of these secondaries
depend both on the gas and CR densities in the cluster, where the CR density
roughly traces the gas outside the core regime and is slightly enhanced in the
center. This density scaling implies that clusters are great targets for
Cherenkov telescopes with a high sensitivity for the central parts of nearby
clusters. Detecting the cluster gamma-ray emission is crucial in this respect as
it potentially provides the unique and unambiguous evidence of CR populations in
clusters through observing the $\pi^0$ bump at about 100~MeV in the spectra.

We adopt the universal spectral and spatial gamma-ray model developed by Pinzke
\& Pfrommer \cite{2010MNRAS.409..449P} to estimate the emission from decaying
$\pi^0$:s that dominates over the IC emission from primary and secondary
electrons above 100~MeV in clusters. The gamma-ray formalism was derived from
high-resolution simulations of clusters of galaxies that included radiative
hydrodynamics, star formation, supernova feedback, and followed CR physics using
a novel formulation that trace the most important injection and loss processes
self-consistently while accounting for the CR pressure in the equation of motion
\cite{2008A&A...481...33J,2007A&A...473...41E,2006MNRAS.367..113P}.  We
highlight two main uncertainties of the models, namely the acceleration
efficiency at formation shocks and CR transport parameters. First, we note that
the overall normalization of the CR and gamma-ray distribution scales with the
maximum acceleration efficiency at structure formation shock waves. Following
recent observations at supernova remnants \cite{2009Sci...325..719H} as well as
theoretical studies \cite{2005ApJ...620...44K}, we assume the maximally allowed
acceleration efficiency for {\em strong} shocks that transfers 50\% of the
shock-dissipated energy (kinetic energy corrected for adiabatic compressional
heating at the shock) to CRs. These efficiencies drop quickly for weaker shocks
\cite{2007A&A...473...41E} which dominate the gravitational energy dissipation
inside galaxy clusters \cite{2006MNRAS.367..113P}. These efficiencies will have
to be corrected down if this value is not realized at strong structure formation
shocks. Second, these simulations (and by extension the analytic model) neglect
active CR transport such as streaming and diffusion relative to the gas, i.e.,
we assume that advective transport dominates and CRs are tightly coupled to the
gas via magnetic fields tangled on sufficiently small scales which produces
centrally enhanced profiles. However, CR diffusion and streaming tends to drive
the CR radial profiles towards being flat, with equal CR number density
everywhere. While the CR streaming velocity is usually larger than typical
advection velocities and becomes comparable or lower than this only for periods
with trans- and supersonic cluster turbulence during a cluster merger. As a
consequence a bimodality of the CR spatial distribution may result with merging
(relaxed) clusters showing a centrally concentrated (flat) CR energy density
profile \cite{2011A&A...527A..99E}. This translates into a bimodality of the
expected diffuse radio and gamma-ray emission of clusters, since more centrally
concentrated CR will find higher target densities for hadronic CR proton
interactions \cite{2011A&A...527A..99E}. As a result of this, relaxed clusters
could have a reduced gamma-ray luminosity by up to a factor of five.

@@

We focus in this section on the spectral distribution of gamma-rays
from clusters. In the LP model, DM annihilation radiation includes
final state radiation and gives rise to substantial amounts of
electrons and positrons that IC upscatter background radiation fields
to high energies. We also consider four supersymmetric DM models with
a high gamma-ray yield in the form of continuum emission and IC
induced emission. In addition to the annihilating DM, we estimate the
gamma-ray flux induced by shock-accelerated CRs. \C{Note that all
  fluxes in this section are derived within an angle corresponding to
  $\rvir$ and are convolved with a PSF of $0.1\deg$.}
